\section{Monitors}

We consider monitors with the following operations:

\begin{lstlisting}
cond cv;
wait(cv);
signal(cv);
signal_all(cv);
\end{lstlisting}

\subsection{Problem 10: Barbershop monitor with priorities}

Use the monitor operations listed above
to make a monitor solution to the priority customer barber shop of Exercise 8. Provide your
solution by completing the code below and explain briefly how your solution gives priority to
express customers. \\


\textbf{Solution}

The solution uses Signal and Continue (SC). We make sure to give priority to express customers by signaling them before regular customers.
The barber will always prioritize express customers over regular customers. 

\begin{lstlisting}
monitor Barbershop {
    int seatsInBarbershop = n;
    int nr = 0; // number of regular customers
    int np = 0; // number of priority customers

    // Queue of regular customer
    cond queuer; 

    // Queueu of express customer
    cond queuep; 

    // Signal to wake up barber 
    cond customerReady;
    
    procedure barber(){
        while (true) {
            // Check if waiting customer 
            while(nr + np) == 0 { 
                wait(customerReady)
            }

            // There is a customer
            // Signal express customer before regular.
            if (np > 0) {
                signal(queuep); 
            } else {
                signal(queuer); 
            }
        }
    }
    procedure regularCustomer() {
        while (true) {
            if (nr + np) < seatsInBarbershop {
                nr := nr + 1;
                signal(customerReady);
                wait(queuer);
                nr := nr - 1;
            }
        }
    }
    procedure priorityCustomer() {
        while (true) {
            if (nr + np) < seatsInBarbershop {
                np := np + 1;
                signal(customerReady);
                wait(queuep);
                np := np - 1;
            }
        }
    }
}
\end{lstlisting}

\subsection{Problem 11: Monitor invariant}

What could be a monitor invariant for the Barbershop monitor? 
Explain briefly why the monitor invariant holds for your monitor solution. \\

\textbf{Solution}

The monitor must hold the following requirements: 
\begin{enumerate}
    \item The number of total customers must be less than seats available. 
    \item There must be 0 or more priority customers. The same must hold for regular customers. 
\end{enumerate}

It can be expressed as the following invariant: 

\[    (\text{np} + \text{nr} \leq \text{freeSeats}) \land (\text{np} \geq 0) \land (\text{nr} \geq 0) \]